\documentclass[10pt,aps,pre,twocolumn,nofootinbib,superscriptaddress,longbibliography]{revtex4-1}
\usepackage[colorinlistoftodos,textsize=tiny]{todonotes}
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{color}
\usepackage{hyphenat}
\usepackage{natbib}
\usepackage{microtype}
\usepackage{mathpazo}
\usepackage{cleveref}
%\usepackage[labelformat = empty,position=top]{subcaption}
\usepackage{pgfplots}
\usepackage{pgfgantt}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{tikz}
\usepackage{mathtools}

\SetKwInput{KwInput}{Input}                % Set the Input
\SetKwInput{KwOutput}{Output}              % set the Output

\usetikzlibrary{calc}
\pgfplotsset{compat=newest}
\pgfplotsset{plot coordinates/math parser=false}
\linespread{1.09}

\newcommand{\brho}{\boldsymbol \rho}
\newcommand{\bsigma}{\boldsymbol \sigma}
\newcommand{\btheta}{\boldsymbol \theta}
\newcommand{\bmu}{\boldsymbol \mu}
\newcommand{\bA}{\mathbf{A}}
\newcommand{\bS}{\mathbf{S}}
\newcommand{\bC}{\mathbf{C}}
\newcommand{\bD}{\mathbf{D}}
\newcommand{\bE}[1]{\operatorname{\mathbb{E}}\left\lbrack #1\right\rbrack}
%\newcommand{\bE}[1]{\left \langle #1 \right\rangle}
\newcommand{\bH}{\mathbf{H}}
\newcommand{\bI}{\mathbf{I}}
\newcommand{\bL}{\mathbf{L}}
\newcommand{\bMc}{\mathbfcal{M}}
\newcommand{\bM}{\mathbf{M}}
\newcommand{\bN}{\mathbf{N}}
\newcommand{\bT}{\mathbf{T}}
\newcommand{\bW}{\mathbf{W}}
\newcommand{\bXc}{\mathbfcal{X}}
\newcommand{\bz}{\mathbf{z}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\bY}{\mathbf{Y}}
\newcommand{\Tr}[1]{\operatorname{Tr}\left\lbrack #1\right\rbrack}
%\newcommand{\bE}[1]{\left\langle #1 \bigl \vert  \btheta \right\rangle }
\DeclareMathOperator*{\argmax}{argmax}
\newcommand{\argmin}{\operatorname{argmin}}
\renewcommand{\Im}[1]{\operatorname{Im} #1 }
\newcommand{\bra}[1]{\left \langle #1 \right|}
\newcommand{\ket}[1]{\left | #1 \right \rangle}

\Crefname{equation}{Eq.}{Eqs.}
\pgfdeclarelayer{background}
\pgfdeclarelayer{foreground}
\pgfsetlayers{background,main,foreground}
\usepackage{pgfplots}
\pgfplotsset{compat=newest}

\title{Thermodynamic Interpretation and Numerical Verification of Wilson's Algorithm for Uniform Spanning Trees}
\author{Carlo Nicolini}
\affiliation{Center for Neuroscience and Cognitive Systems, Istituto Italiano di Tecnologia}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
We revisit Wilson's algorithm for sampling uniform spanning trees and its
intimate connection with effective resistance. We present numerical
experiments across representative graph families that validate the identity
\(p_e = w_e R_e\) where $p_e$ is the marginal probability that an undirected
edge $e$ belongs to a uniformly sampled spanning tree, $w_e$ its weight and
$R_e$ the effective resistance between its endpoints. Our experiments compare
deterministic linear-algebraic computation of marginals and Monte Carlo
estimates via Wilson sampling, and study solver trade-offs for scaling to
larger graphs.
\end{abstract}

\section{Introduction}
Wilson's algorithm is a classical procedure to sample a uniform spanning tree
(UST) of a connected graph using loop-erased random walks. A celebrated
identity relates UST edge marginals to effective resistance: $p_e = w_e R_e$.
This work provides a concise numerical study verifying the identity across
graph families and documents a reproducible pipeline for generating the
reported metrics and figures.

\section{Background and Theory}
We recall a few definitions. Given a connected, undirected graph $G=(V,E)$
with nonnegative edge weights $w_e$, the Laplacian $L$ is defined by
$L_{ii} = \sum_j w_{ij}$ and $L_{ij} = -w_{ij}$ for $i\ne j$. For a grounded
Laplacian $\tilde L$ (obtained by removing the row and column of a chosen
root vertex), the effective resistance between two vertices $u,v$ can be
computed by solving $\tilde L x = b$ where $b$ encodes $+1$ and $-1$ at the
reduced coordinates of $u$ and $v$, and then forming $R_{uv} = b^\top x$.
The UST marginal identity reads $p_{uv} = w_{uv} R_{uv}$.

The parameter $q$ appearing in rooted forest models connects to spectral
quantities: the expectation of the number of roots for the $q$-rooted forest
distribution equals $s(q) = q\,\mathrm{Tr}((qI + L)^{-1}) = \sum_i q/(q+\lambda_i)$,
which we compute in the experiments and expose as a fitted attribute in our
estimator.

\section{Methods}
We implemented:
\begin{itemize}
  \item Graph generators: ER, BA, and grid graphs.
  \item A scikit-learn-like Wilson estimator (\texttt{src/wilson/wilson.py})
    exposing a \texttt{fit(G)} method and \texttt{s_} when $q$ is provided, and
    a \texttt{sample(seed)} method for Monte Carlo sampling.
  \item An experiments runner (\texttt{scripts/run_wilson_experiments.py}) that
    computes dense effective-resistance marginals (via pseudoinverse), runs
    repeated Wilson sampling to estimate empirical marginals, and saves CSVs
    and figures into \texttt{artifacts/}.
\end{itemize}

\section{Experiments}
We executed the pipeline on representative graphs. For each graph we report
the Pearson correlation and RMSE between the theoretical marginals and the
empirical frequencies from Wilson sampling. Recoverable artifacts (per-edge
CSVs and scatter plots) are written to the repository's \texttt{artifacts/}
folder for reproducibility.

% Include generated results fragment if present
\IfFileExists{docs/results_generated.tex}{\input{docs/results_generated.tex}}{%
  \input{docs/results.tex}
}

\section{Discussion}
The experiments show very good agreement between Wilson sampling and the
effective-resistance formula across the tested graph families. The Wilson
sampler provides a lightweight Monte Carlo approach that complements
deterministic linear solvers when scaling beyond moderate-sized graphs.

\section{Reproducibility}
Run:\\
\begin{verbatim}
python scripts/run_wilson_experiments.py
python scripts/generate_results_tex.py
latexrun paper.tex
\end{verbatim}

\section{Conclusion}
We provided numerical validation of the $p_e = w_e R_e$ identity and
packaged the experiments and figure generation into reproducible scripts.

\bibliographystyle{apsrev4-1}
\bibliography{biblio3}

\end{document}
